The Efficient Handling of Guards \\in the Design of RPython's Tracing JIT

### 2012-11-09 talk/vmil2012/paper.tex
typos
Start preparing submission for publication
do not require pdf for figures
force page size to US Letter
Conditionally use hyperref or url packages depending on the target

\documentclass[10pt,preprint]{sigplanconf}

\usepackage{ifthen}
\usepackage{sparklines}
\usepackage{booktabs}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{colortbl}
\usepackage{wrapfig}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{relsize}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage[pdfpagelabels=true]{hyperref}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{flafter}
\usepackage{enumitem}
\setitemize{noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt}

@@

\usepackage[T1]{fontenc}
\usepackage[scaled=0.81]{beramono}

@@

\hypersetup{%
  plainpages=false,%
  %hyperfootnotes=false,%
  colorlinks=true,%
  urlcolor=black,%
  citecolor=black,%
  linkcolor=black,%
  pdftitle={The Efficient Handling of Guards in the Design of RPython's Tracing JIT},%
  pdfauthor={David Schneider},
}

@@

\conferenceinfo{VMIL'12}{}
\CopyrightYear{2012}
\crdata{}

@@

There are several forces guiding the design of resume data handling.
Guards are a very common operations in the traces.
However, as will be shown, a large percentage of all operations
are optimized away before code generation.
Since there are a lot of guards
the resume data needs to be stored in a very compact way.
On the other hand, tracing should be as fast as possible,
so the construction of resume data must not take too much time.

@@

\begin{figure}[ht]
\includegraphics[width=0.4\textwidth]{figures/resume_data.pdf}
\vspace{-3mm}
\caption{The resume data for Figure~\ref{fig:trace-log}}
\label{fig:resume-data}
\end{figure}

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split.
The guard becomes the branching point of two conditional paths with no
additional overhead.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard~\#1 jumps to the trampoline, loads the backend map and
then calls the bailout handler, whereas Guard~\#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

\subsection{Guard Failures}
\label{sub:guard_failure}
The last point in this discussion is the frequency of guard failures.
Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
relative amounts of guards that ever fail and of guards that fail often enough that a bridge is compiled.\footnote{
    The threshold used is 200 failures. This rather high threshold was picked experimentally to give
    good results for long-running programs.
}
It also contains sparklines depicting the failure rates for the failing guards
in decreasing order, each normalized to the most failing guard.
The numbers presented for guards that have a bridge represent the
failures up to the compilation of the bridge and all executions of the then
attached bridge.

>>>>

\documentclass[letter,10pt]{sigplanconf}
\special{papersize=8.5in,11in}

\usepackage{ifthen}
\usepackage{sparklines}
\usepackage{booktabs}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{colortbl}
\usepackage{wrapfig}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{relsize}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{flafter}
\usepackage{enumitem}
\setitemize{noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt}

@@

\usepackage[T1]{fontenc}
\usepackage[scaled=0.81]{beramono}
\usepackage{ifpdf}

@@

\ifpdf
\usepackage[pdfpagelabels=true,hidelinks]{hyperref}
\hypersetup{%
  plainpages=false,%
  pdftitle={The Efficient Handling of Guards in the Design of RPython's Tracing JIT},%
  pdfauthor={David Schneider},
}
\else
\usepackage[hyphens]{url}
\fi

@@

\conferenceinfo{VMIL'12,} {October 21, 2012, Tucson, Arizona, USA.}
\CopyrightYear{2012}
\copyrightdata{978-1-4503-1633-0/12/10}
\crdata{}

@@

There are several forces guiding the design of resume data handling.
Guards are a very common operation in the traces.
However, as will be shown, a large percentage of all operations
are optimized away before code generation.
Since there are a lot of guards
the resume data needs to be stored in a very compact way.
On the other hand, tracing should be as fast as possible,
so the construction of resume data must not take too much time.

@@

\begin{figure}[ht]
\includegraphics[width=0.4\textwidth]{figures/resume_data}
\vspace{-3mm}
\caption{The resume data for Figure~\ref{fig:trace-log}}
\label{fig:resume-data}
\end{figure}

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split.
The guard becomes the branching point of two conditional paths with no
additional overhead.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard~\#1 jumps to the trampoline, loads the backend map and
then calls the bailout handler, whereas Guard~\#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

\subsection{Guard Failures}
\label{sub:guard_failure}
The next point in this discussion is the frequency of guard failures.
Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
relative amounts of guards that ever fail and of guards that fail often enough that a bridge is compiled.\footnote{
    The threshold used is 200 failures. This rather high threshold was picked experimentally to give
    good results for long-running programs.
}
It also contains sparklines depicting the failure rates for the failing guards
in decreasing order, each normalized to the most failing guard.
The numbers presented for guards that have a bridge represent the
failures up to the compilation of the bridge and all executions of the then
attached bridge.
